Bias-free model fitting of correlated data in interferometry

### 2021-04-09 article.tex
revision

\renewcommand{\floatpagefraction}{.9}%

@@

\begin{abstract}
In optical and infrared long-baseline interferometry, data often display significant correlated errors because of uncertain multiplicative factors such as the instrumental transfer function or the pixel-to-visibility matrix.  In the context of model fitting, this situation often leads to a significant bias. In the gravest cases can show a fit passing outside of all error bars. It is known in nuclear physics as Peelle's Pertinent Puzzle.  I show how this arises in the context of interferometry and how severe it can be in a realistic setting.  I then give a conceptually simple and computationally cheap way to avoid the issue: model the data without covariances, estimate the covariance matrix by error propagation using the modelled data instead of the actual data, and perform the model fitting using the covariance matrix.
\end{abstract}

@@

In spite of strong evidence of correlations in data, due to redundancy \citep[][in the case of closure phases]{MON07}, calibration \citep{PER03}, or atmospheric biases acting on all spectral channels in the same way \citep{LAW00}, only a few authors \citep{PER04,ABS06,BER06,LAC19,KAM20} have analysed correlated data while most assume independent errors.  In particular, the only interferometric instrument I know of with a data processing software taking into account one source of correlations---calibration---is FLUOR\footnote{Fiber Linked Unit for Optical Recombination} \citep[at IOTA\footnote{Infrared and Optical Telescope Array}, then CHARA\footnote{Center for High Angular Resolution Array}, ][]{PER04}. None of the five first and second-generation ones at the VLTI\footnote{Very Large Telescope Interforometer} does \citep{AMBER,MIDI,PIONIER,GRAVITYpipe,MATISSEpipe}.  The same lack of support for correlations is present in image reconstruction programmes \citep[e.g. MIRA, see][]{THI08}, model-fitting tools \citep[e.g. Litpro, see][]{TAL08}, or the still widespread first version of the Optical Interferometric FITS format \citep[OIFITS v. 1,][]{OIFITS1}.

Unfortunately, ignoring uncertainties may lead to significant errors in model parameters as \citet{LAC19} evidenced with stellar diameters using PIONIER\footnote{Precision Integrated-Optics Near-infrared Imaging ExpeRiment} \citep{PIONIER} data at the VLTI. Also \citet{KAM20} established that correlations were necessary to achieve a higher contrast ratio in companion detection using GRAVITY \citep{GRAVITY} at the VLTI.

Several sources of correlated uncertainties occur in a multiplicative context, when several data points are normalised with the transfer function \citep{PER03} or the coherent fluxes are derived with the pixel-to-visibility matrix formalism \citep{TAT07}.  In both cases, the uncertainty on the multiplicative factor translates into a systematic, correlated one in the final data product. In the context of experimental nuclear physics, \citet{PEE87} noted that this scenario could lead to an estimate falling below the individual data points, a paradox known as Peelle's Pertinent Puzzle (PPP). It results from the usual and straightforward, but actually incorrect, way to propagate covariances \citep{DAG94,NEU12}.  A few workarounds have been proposed \citep[e.g.][]{BUR11,BEC12,NIS14} but they are not straightforward.

@@

\section{Peelle's Pertinent Puzzle basics}
\label{sec:ppp}

I rewrite and adapt Peelle's original example in the context of long-baseline interferometry. Let's assume the inverse of the instrumental fringe contrast $\cot \pm \cot\reldev$ has been interpolated from one or several calibrator observations. A visibility amplitude is now now estimated from two contrast measurements $\raw_1 \pm \absdev$ and $\raw_2 \pm \absdev$.  For each measurement, visibility amplitudes are:

\begin{subequations}
\begin{align}
    \data_1 &= \system{\cot\raw_1}{\cot\absdev}{\cot\raw_1\reldev},\\
    \data_2 &= \system{\cot\raw_2}{\cot\absdev}{\cot\raw_2\reldev}.
\end{align}
\label{eq:data12}
\end{subequations}

They are normalised with the same quantity (\cot), so they are correlated, hence the systematic uncertainty term between parentheses in \eqref{data12}. Error propagation gives the covariance matrix

\begin{equation} 
   \vcov = \begin{pmatrix} 
     \absdev^2\cot^2 + \reldev^2\data_1^2 & \reldev^2\data_1\data_2\\
     \reldev^2\data_1\data_2              & \absdev^2\cot^2 + \reldev^2\data_2^2
            \end{pmatrix},
\end{equation}
when the second-order error term ($\absdev\reldev$) is ignored.  Using the matrix inverse $\vcov^{-1}$, the least squares estimate is 
\begin{equation}
\begin{split}
    {\data} &= \frac{  (\vcov^{-1}_{11} + \vcov^{-1}_{12}) \data_1
                      +(\vcov^{-1}_{22} + \vcov^{-1}_{12}) \data_2 }
                    {   (\vcov^{-1}_{11} + \vcov^{-1}_{12}) \hphantom{\data_1}
                      + (\vcov^{-1}_{22} + \vcov^{-1}_{12}) \hphantom{\data_2} 
                    },\\
            &= \frac{\data_1 + \data_2}{2} 
            \left(1 + \reldev^2\frac{(\data_1-\data_2)^2}{2\tau^2\absdev^2}\right)^{-1}.
\end{split}
\end{equation}
It systematically falls below the average of the two values $\data_1$ and $\data_2$.  If measurements differ significantly, it can even fall below the lowest value.  Figure~\ref{fig:peelle} gives such an example with a instrumental visibility of 50\% and two measurements on an unresolved target: 
\begin{align*}
    \cot   &= 2.000 \pm 0.100\ (5\%),\\
    \raw_1 &= 0.495 \pm 0.003\ (\approx 0.6\%),\\
    \raw_2 &= 0.505 \pm 0.003\ (\approx 0.6\%),
\intertext{which yields two points 2.4 standard deviations apart}
    \data_1 &= \system{0.990}{0.006}{0.050},\\
    \data_2 &= \system{1.010}{0.006}{0.050}
\intertext{and the estimate}
    \data &= \system{0.986}{0.004}{0.049}
\end{align*}
falls outside the data range.
\begin{figure}
\includegraphics[width=\linewidth]{pdf/original-peelle.pdf}
\caption{Original Peelle problem rewritten in the context of interferometry.  \emph{Bottom:} Two raw visibility amplitudes $\nu_1$ and $\nu_2$ (points with statistic error bars of $\approx 0.6\%$) are calibrated by the transfer function $1/\tau$ (solid line with systematic error zone of 5\%). \emph{Top:} the two calibrated visibility measurements $\data_1$ and $\data_2$ (points with statistic error bars of $\approx 0.6\%$) are strongly correlated. The least-squares estimate for the visibility $\data$ (dashed line, with statistic uncertainty error zone displayed) falls outside of the data range.}
\label{fig:peelle}
\end{figure}

\begin{table}
\caption{Symbols used in this paper. Lower case bold font is used for vectors and upper case bold font for matrices.}
\begin{tabular}{ll}
\hline\hline
Symbol                        & Meaning\\
\hline
$\mean a$                     & true value of $a$\\
$<a>$                         & expectancy of $a$\\
$\tr{\vec A}$                 & transpose of $\vec A$\\
$\vec c = \vec a\hadam\vec b$ & element-wise product of $\vec a$ and $\vec b$\\
$\vec C = \vec a\outer\vec b$ & outer product of $\vec a$ and $\vec b$\\
$\vec c = \vec A\vec b$       & matrix product of $\vec A$ and $\vec b$\\
$\delta$                      & Kronecker delta\\
\hline
$\vdata$                      & data\\ 
$\verror$                     & error ($= \vdata - \vdatamean$)\\
$\dev^2$                      & variance ($= \expect{\error^2}$)\\
$\vcov$                       & covariance matrix ($= \expect{\verror\outer\verror}$)\\
$\corr$                       & correlation coefficient\\
$\vsens$, $\msens$            & sensitivity vector or matrix\\
$\vparam$                     & parameters of the model\\
$\vmod$                       & model values ($= \msens\vparam\approx\vdatamean$)\\
\hline
$a\meas$                      & $a$ of the measurement error\\
$a\norm$                      & $a$ of the normalisation error\\
$\ppp a$                      & $a$ impacted by PPP\\
\hline
\end{tabular}
\end{table}

\section{Single value fit}
\label{sec:single}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{pdf/uncorrelated-peelle.pdf}
\caption{Fit $\datappp$  to unresolved visibilities ($\data = 1$), as a function of the relative uncertainty on the calibration $\reldev$ and the number of measurements $n$. $2/n\times10^5$ simulations were made and averaged, assuming that $\abserror$ and $\relerror$ follow normal distributions. \textit{Top:} fully correlated normalisation like in original Peelle's puzzle ($\absdev=0.02$ and $\corr = 1$). \textit{Bottom:} normalisation error without correlation ($\absdev=0.02$ and $\corr = 0$).}
\label{fig:uncorr-peelle}
\end{figure}

Let's consider a set of measurements of a single normalised quantity, like the visibility amplitude, given by a column vector $\vdata = \tr{(\data_1, \cdots, \data_n)}$. It is derived from an uncalibrated quantity like the fringe contrast, $\vraw = \tr{(\raw_1, \cdots, \raw_n)}$ and a normalisation factor, like the cotransfer function, $\vcot = \tr{(\cot_1, \cdots, \cot_n)}$ by 
\begin{equation}
    \vdata = \vcot\hadam\vraw,
\end{equation}
where $\hadam$ denotes the Hadamard (element-wise) product of vectors. With $\datamean$, $\cotmean$, and $\rawmean$ the true, but unknown, values of these quantities, the error vector on $\vdata$  
\begin{align}
    \verror    &= \vdata - \datamean, \\
\intertext{can be written as a sum of measurement and normalisation errors if one ignores a second-order term:}
    \verror    &= \vabserror + \datamean \vrelerror.\\
\intertext{These errors are given by}
    \vabserror &= \cotmean (\vraw - \rawmean), \\
    \vrelerror &= \frac1\cotmean (\vcot-\cotmean).
\end{align}

Let's assume $\vabserror$ and $\vrelerror$ are independent error terms of mean 0 and standard deviations are $\absdev$ and $\reldev$, respectively. In addition, I consider correlation of the normalisation errors, with correlation coefficient $\corr$.  In the case of interferometry, it can arise from the uncertainty on the calibrators' geometry.  The covariance matrix is given by
\begin{align}
    \vcov     &= \expect{\verror\outer\verror},\\
\intertext{where $\outer$ denotes the outer product of vectors and $\expect{}$ stands for the expectancy, so that}
    \cov_{ij} &= \left[ \absdev^2 
                  + (1-\corr)\reldev^2\datamean^2
                \right] \delta_{ij} 
            + \corr\reldev^2\datamean^2.
\end{align}

The value $\datamean$ is yet to be determined, so the covariances are often derived using the measurements $\vdata$ in the propagation:
\begin{equation}
    \ppp{\cov_{ij}} = \left[ 
                    \absdev^2 
                  + (1-\corr)\reldev^2 \data_i^2
                \right] \delta_{ij} 
            + \corr\reldev^2\data_i\data_j. \label{eq:cov}
\end{equation}

The least squares estimate for $\datamean$ is given by
\begin{equation}
    \ppp{\mod} =
                  \frac { \tr\vsens{\ppp\vcov}^{-1}\vdata }
                        { \tr\vsens{\ppp\vcov}^{-1}\vsens }
\end{equation}
where $\vsens = \tr{(1, \cdots, 1)}$ is the trivial sensitivity vector.  The covariance matrix is the sum of an invertible diagonal matrix and one of rank one---see \eqref{cov}---, so that the inverse is obtained using the Woodbury matrix identity: 
\begin{align}
    \{ {\ppp\vcov}^{-1} \}_{ij} &= \frac{\delta_{ij}}{\dev_i^2}
         - \frac{\corr\reldev^2\data_i\data_j}
                {\dev_i^2\dev_j^2\Big(1 + 
            \corr\reldev^2\sum\limits_{k} \frac{\data_k^2}{\dev_k^2}\Big)},\\
  \intertext{where}
  \dev_i^2 &= \absdev^2+(1-\corr)\reldev^2\data_i^2.
\end{align}
It allows me to put the least squares estimate in the relatively compact form 
\begin{equation}
  \ppp{\mod} = 
    \frac { 
      \sum\limits_i \frac{\data_i}{\dev_i^2}
    }{ 
      \sum\limits_i 
          \Big( 1   
             + \corr\reldev^2\sum\limits_{j} \frac{(\data_i-\data_j)^2}{\dev_j^2}
           \Big)
          \frac{1}{\dev_i^2} 
   }.\label{eq:mu}
\end{equation}

For small enough errors ($\error_i \ll \datamean$) the second-order Taylor development in $\error_i = \data_i - \datamean$ yields 
\begin{equation}
    \ppp{\mod} \approx  \datamean 
                  +\frac1n {\sum\limits_i \error_i}
                       - \frac {[2 + (n - 2)\corr]\reldev^2\datamean^2
                          }{
                           2n^2 \devmean^2             
                          }
                           \sum\limits_{i\ne j} (\error_i-\error_j)^2
\end{equation}
where
\begin{equation}
    \devmean^2 = \absdev^2+(1-\corr)\reldev^2\datamean^2 
\end{equation}

Since $\expect{\error_i} = 0$ and $\expect{(\error_i-\error_j)^2} = 2\devmean^2$, the expectancy
\begin{equation}
    \expect{\ppp{\mod}} \approx \datamean 
            \left[ 1 -  
             \left(1-\frac1n\right)
             \left(2 + (n-2)\corr\right)
              \reldev^2 \right]
\end{equation}
is biased.  If the data are not correlated ($\corr = 0$), the bias is small
($\reldev^2$ to $2\reldev^2$) but it becomes large for correlated data if the number of points is large ($\sim n\reldev^2$ for fully correlated data) as \citet{DAG94} already noted. Figure~\ref{fig:uncorr-peelle} shows a simulation of the bias as a function of the normalisation uncertainty $\reldev$ for various data sizes ($n = 2$ to 100). This analytical derivation confirms the numerical simulation by \citet[][see their Fig.~4]{NEU12}.

For spectro-interferometric observations with 4 telescopes, the number of correlated points can be over 1,000, so even with a low correlation coefficient, the impact can be large. For instance, a single GRAVITY observation in high spectral resolution yields $n = 6 \times 210$ visibility amplitudes. With an observed correlation of $\rho \approx 16$\% in the instrumental visibility amplitudes \citep{KAM20} and a typical $\reldev = 1$--2\% normalisation error, the bias on the calibrated visibilities could be 2--8\%.

@@

\begin{figure}
\centering
\subfigure[Under-resolved, linear least squares]{\includegraphics[width=\linewidth]{pdf/fit-example-quadratic.pdf}\label{fig:fitexample}}\\
\subfigure[Well resolved, non-linear least squares]{\includegraphics[width=\linewidth]{pdf/fit-example-gauss.pdf}\label{fig:fitexample2}}
\caption{Example of a fit of correlated data from a four telescope interferometer with medium resolution: 6 groups of 100 data points (light gray points with the measurement error bar) with 2\% uncorrelated measurement errors and 3\% correlated normalisation.  \emph{Top:} Simulated under-resolved data $\data = 1-x^2$ (thick gray line) are fitted with linear least-squares model $\mod = a-bx^2$ using the four prescriptions for the covariance matrix. \emph{Bottom:} The same for well-resolved data $\data = \exp -3x^2$ and non-linear least-squares with model $\mod = a\exp -bx^2$.}
\end{figure}

>>>>

\renewcommand{\floatpagefraction}{.9}%
\def\correction#1{{\bfseries #1}}

@@

\begin{abstract}
In optical and infrared long-baseline interferometry, data often display significant correlated errors because of uncertain multiplicative factors such as the instrumental transfer function or the pixel-to-visibility matrix \correction{in the model parameters}.  In the context of model fitting, this situation often leads to a significant bias. In the \correction{most severe this can can result in a fit lying outside of the range of measurement values. This} is known in nuclear physics as Peelle's Pertinent Puzzle.  I show how this arises in the context of interferometry and \correction{determine that the relative bias is of the order of the square root of the correlated component of the relative uncertainty times the number of measurements. It impacts preferentially large data sets, such as those obtained in medium to high spectral resolution}.  I then give a conceptually simple and computationally cheap way to avoid the issue: model the data without covariances, estimate the covariance matrix by error propagation using the modelled data instead of the actual data, and perform the model fitting using the covariance matrix. \correction{I also show that a more imprecise but also unbiased result can be obtained from ignoring correlations in the model fitting.}
\end{abstract}

@@

In spite of strong evidence of correlations in \correction{the} data, due to redundancy \citep[][in the case of closure phases]{MON07}, calibration \citep{PER03}, or atmospheric biases acting on all spectral channels in the same way \citep{LAW00}, only a few authors \citep{PER04,ABS06,BER06,LAC19,KAM20} have \correction{accounted for these correlations} while most assume\correction{d statistically} independent errors.  In particular, the only interferometric instrument I know of with a data processing software taking into account one source of correlations---calibration---is FLUOR\footnote{Fiber Linked Unit for Optical Recombination} \citep[at IOTA\footnote{Infrared and Optical Telescope Array}, then CHARA\footnote{Center for High Angular Resolution Array}, ][]{PER04}. None of the five first and second-generation ones at the VLTI\footnote{Very Large Telescope Interforometer} does \citep{AMBER,MIDI,PIONIER,GRAVITYpipe,MATISSEpipe}.  The same lack of support for correlations is present in image reconstruction programmes \citep[e.g. MIRA, see][]{THI08}, model-fitting tools \citep[e.g. Litpro, see][]{TAL08}, or the still widespread first version of the Optical Interferometric FITS format \citep[OIFITS v. 1,][]{OIFITS1}.

Unfortunately, ignoring \correction{correlations} may lead to significant errors in model parameters as \citet{LAC19} evidenced with stellar diameters using PIONIER\footnote{Precision Integrated-Optics Near-infrared Imaging ExpeRiment} \citep{PIONIER} data at the VLTI. Also \citet{KAM20} established that \correction{accounting for} correlations is necessary to achieve a higher contrast ratio in companion detection using GRAVITY \citep{GRAVITY} at the VLTI.

Several sources of correlated uncertainties occur in a multiplicative context, when several data points are normalised with the transfer function \citep{PER03} or the coherent fluxes are derived with the pixel-to-visibility matrix formalism \citep{TAT07}.  In both cases, the uncertainty on the multiplicative factor translates into a systematic, correlated one in the final data product. In the context of experimental nuclear physics, \citet{PEE87} noted that this scenario could lead to an estimate falling below the individual data points, a paradox known as Peelle's Pertinent Puzzle (PPP). It results from the usual, but actually incorrect, way to propagate covariances, \correction{in which the measured values are used in the calculations} \citep{DAG94,NEU12}.  A few workarounds have been proposed but they are \correction{either computationally expensive \citep[e.g. sampling of the posterior probability distribution for Bayesian analysis, see][]{NEU12} or require a conceptually difficult implementation \citep{BEC12,NIS14}}.

@@

\section{Peelle's Pertinent Puzzle}
\label{sec:ppp}
I rewrite and adapt Peelle's original example in the context of long-baseline interferometry \correction{\citep[see][Sect.~1 \& 2.1]{NEU12}}. \correction{One or several calibrator observations yield the inverse of the instrumental fringe contrast $\cot \pm \cot\reldev$.  I use the relative uncertainty $\reldev$ on the transfer function as it is often referred to in percentage terms.} A visibility amplitude is now estimated from two contrast measurements $\raw_1 \pm \absdev$ and $\raw_2 \pm \absdev$.  For each measurement, the visibility amplitudes are:

\begin{subequations}
\begin{align}
    \data_1 &= \system{\cot\raw_1}{\cot\absdev}{\cot\raw_1\reldev},\\
    \data_2 &= \system{\cot\raw_2}{\cot\absdev}{\cot\raw_2\reldev},
\end{align}
\label{eq:data12}
\end{subequations}
\correction{where the second-order error term $\cot\reldev\absdev$ has been ignored}.

They are normalised with the same quantity (\cot), so they are correlated, hence the systematic uncertainty term between parentheses in \eqref{data12}. Error propagation \correction{yields} the covariance matrix

\begin{figure}
\includegraphics[width=\linewidth]{pdf/original-peelle.pdf}
\caption{Original Peelle problem rewritten in the context of interferometry.  \emph{Top:} Two raw visibility amplitudes $\nu_1$ and $\nu_2$ (points with statistic error bars of $\approx 0.6\%$) are calibrated by the transfer function $1/\tau$ (solid line with systematic error zone of 5\%). \emph{Bottom:} the two calibrated visibility measurements $\data_1$ and $\data_2$ (points with statistic error bars of $\approx 0.6\%$) are strongly correlated. The least-squares estimate for the visibility $\data$ (dashed line, with statistic uncertainty error zone displayed) falls outside of the data range.  The systematic error on $\data$, $\data_1$, and $\data_2$ is shown on the right.}
\label{fig:peelle}
\end{figure}

\begin{equation} 
   \vcov = \begin{pmatrix} 
     \absdev^2\cot^2 + \reldev^2\data_1^2 & \reldev^2\data_1\data_2\\
     \reldev^2\data_1\data_2              & \absdev^2\cot^2 + \reldev^2\data_2^2
            \end{pmatrix}.
\end{equation}
\correction{Under the hypothesis of Gaussian errors, I obtain} the least squares estimate using the weight matrix $\vec{W} = \vcov^{-1}$:
\begin{gather}
\begin{split}
    {\data} &= 
              %  \frac{  (\vcov^{-1}_{11} + \vcov^{-1}_{12}) \data_1
              %        +(\vcov^{-1}_{22} + \vcov^{-1}_{12}) \data_2 }
              %      {   (\vcov^{-1}_{11} + \vcov^{-1}_{12}) \hphantom{\data_1}
              %        + (\vcov^{-1}_{22} + \vcov^{-1}_{12}) \hphantom{\data_2} 
              %      },
                \frac{  \data_1              W_{11} 
                      + (\data_1 + \data_2)  W_{12} 
                      + \data_2              W_{22} 
                     }
                    {   W_{11} 
                      + 2 W_{12} 
                      + W_{22}  
                    },
                 \\
            &= \frac{\data_1 + \data_2}{2} 
            \left(1 + \reldev^2\frac{(\data_1-\data_2)^2}{2\tau^2\absdev^2}\right)^{-1},
\end{split} \label{eq:mu2}\\
\intertext{\correction{with the uncertainty}}
\begin{split}
    \datadev^2 &= \frac{1}{W_{11} + 2W_{12} 
                     + W_{22}},\\
               &=   \left[ 
                        \left(\frac{\tau\absdev}{\sqrt{2}}\right)^2 
                      +  \frac{\data_1^2 + \data_2^2}{2}\reldev^2 
                    \right]
                    \frac{2\data}{\data_1 + \data_2}. \label{eq:datadev}
\end{split} 
\end{gather}
\correction{The visibility estimate $\data$} systematically falls below the average of the two values $\data_1$ and $\data_2$.  If \correction{the} measurements differ significantly, it can even fall below the lowest value.  Figure~\ref{fig:peelle} gives such an example with a\correction{n} instrumental visibility of 50\% and two measurements on an unresolved target: 
\begin{align*}
    \cot   &= 2.000 \pm 0.100\ (\reldev = 5\%),\\
    \raw_1 &= 0.495 \pm 0.003\ (\absdev/\raw_1 \approx 0.6\%),\\
    \raw_2 &= 0.505 \pm 0.003\ (\absdev/\raw_2 \approx 0.6\%),
\intertext{which yields two points 2.4 standard deviations apart}
    \data_1 &= \system{0.990}{0.006}{0.051},\\
    \data_2 &= \system{1.010}{0.006}{0.050}
\intertext{and the \correction{visibility amplitude} estimate}
    \data &= \system{0.986}{0.004}{0.050}
\end{align*}
falls outside the data range.  \correction{The uncertainties quoted for $\data$ correspond to the first and second terms within the square brackets of Eq.~(\ref{eq:datadev}).}

\begin{table}
\caption{Symbols used in this paper. Lower case bold font is used for vectors and upper case bold font for matrices.}
\begin{tabular}{ll}
\hline\hline
Symbol                        & Meaning\\
\hline
$\mean a$                     & true value of $a$\\
$<a>$                         & \correction{expected value} of $a$\\
$\tr{\vec A}$                 & transpose of $\vec A$\\
$\vec c = \vec a\hadam\vec b$ & element-wise product of $\vec a$ and $\vec b$\\
$\vec C = \vec a\outer\vec b$ & outer product of $\vec a$ and $\vec b$\\
$\vec c = \vec A\vec b$       & matrix product of $\vec A$ and $\vec b$\\
$\delta$                      & Kronecker delta\\
\hline
$\vdata$                      & data\\ 
$\verror$                     & error ($= \vdata - \vdatamean$)\\
$\rerror$                     & relative error ($= \error / \vdatamean$)\\
$\dev$                        & deviation ($= \sqrt{\expect{\error^2}}$), uncertainty\\
$\rdev$                       & relative uncertainty ($= \sqrt{\expect{\rerror^2}}$)\\
$\vcov$                       & covariance matrix ($= \expect{\verror\outer\verror}$)\\
$\corr$                       & correlation coefficient\\
$\vsens$, $\msens$            & sensitivity vector or matrix\\
$\vparam$                     & parameters of the model\\
$\vmod$                       & model values ($= \msens\vparam\approx\vdatamean$)\\
\hline
$a\meas$                      & $a$ of the measurement error\\
$a\norm$                      & $a$ of the normalisation error\\
$\ppp a$                      & $a$ impacted by PPP\\
$\sysdev$                     & relative systematic uncertainty\\
$\stadev$                     & relative statistical uncertainty\\
\hline
\end{tabular}
\end{table}

\section{Fit by a constant}
\label{sec:single}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{pdf/uncorrelated-peelle.pdf}
\caption{Fit $\datappp$  to unresolved visibilities ($\data = 1$), as a function of the relative uncertainty on the calibration $\reldev$ and the number of measurements $n$. $2/n\times10^5$ simulations were made and averaged, assuming that $\rawerror$ and $\relerror$ follow normal distributions. \textit{Top:} fully correlated normalisation like in original Peelle's puzzle ($\rawdev=0.02$ and $\corr = 1$). \textit{Bottom:} normalisation error without correlation ($\rawdev=0.02$ and $\corr = 0$).}
\label{fig:uncorr-peelle}
\end{figure}

\correction{I now generalise the results of the last section to an arbitrary number of measurements} of a single normalised quantity, \correction{such as} the visibility amplitude \correction{of an unresolved source, which is expected to be constantly one for all interferometric baselines.} Let the column vector $\vdata = \tr{(\data_1, \cdots, \data_n)}$ \correction{contain the $n$ visibility amplitudes}. It is derived from an uncalibrated quantity like the fringe contrast, $\vraw = \tr{(\raw_1, \cdots, \raw_n)}$ and a normalisation factor, like the cotransfer function, $\vcot = \tr{(\cot_1, \cdots, \cot_n)}$ by 
\begin{equation}
    \vdata = \vcot\hadam\vraw,
\end{equation}
where $\hadam$ denotes the Hadamard (element-wise) product of vectors. With $\datamean$, $\cotmean$, and $\rawmean$ the true, but unknown, values of these quantities, the error vector on $\vdata$  
\begin{align}
    \verror    &= \vdata - \datamean, \\
\intertext{can be written as a sum of measurement and normalisation \correction{relative} errors if one ignores \correction{the} second-order terms:}
    \verror    &= (\vrawerror + \vrelerror) \datamean.\\
\intertext{These errors are given by}
    \vrawerror &= \frac1\rawmean (\vraw - \rawmean), \\
    \vrelerror &= \frac1\cotmean (\vcot-\cotmean).
\end{align}

\correction{I} assume $\vrawerror$ and $\vrelerror$ are independent, of mean 0, and have standard deviations $\rawdev$ and $\reldev$, respectively. In addition, I consider correlation of the normalisation errors, with correlation coefficient $\corr$.  In the case of interferometry, it can arise from the uncertainty on the calibrators' geometry.  The covariance matrix \correction{of the visibility amplitudes} is given by
\begin{align}
    \vcov     &= \expect{\verror\outer\verror},\\
\intertext{where $\outer$ denotes the outer product of vectors and $\expect{}$ stands for the \correction{expected value}, so that}
    \cov_{ij} &= \underbrace{\left[ \rawdev^2 
                  + (1-\corr)\reldev^2 \right]}_{\stadev^2} \datamean^2
                \delta_{ij} 
            + \underbrace{\corr\reldev^2}_{\sysdev^2} \datamean^2.
\end{align}
The non-diagonal diagonal elements of the matrix feature the systematic
relative uncertainty $\sysdev$, i.e. the correlated component of the
uncertainties.  In the case of a fully correlated transfer function ($\corr =
1$), it is equal its uncertainty ($\sysdev = \reldev$). The diagonal term of
the matrix additionally includes the statistical relative uncertainty
$\stadev$, i.e. the uncorrelated component of the uncertainties.  In the case
of a fully correlated transfer function, it is equal to the uncertainty of the
uncalibrated visibility ($\stadev = \rawdev$).

The value $\datamean$ is yet to be determined, so the covariances are often derived using the measurements $\vdata$ in the propagation:
\begin{equation}
    \ppp{\cov_{ij}} = \stadev^2 \data_i^2 \delta_{ij} 
            + \sysdev^2\data_i\data_j. \label{eq:cov}
\end{equation}

The least squares estimate for $\datamean$ is given by
\begin{equation}
    \ppp{\mod} =
                  \frac { \tr\vsens{\ppp\vcov}^{-1}\vdata }
                        { \tr\vsens{\ppp\vcov}^{-1}\vsens }
    \label{eq:pppmod}
\end{equation}
where $\vsens = \tr{(1, \cdots, 1)}$ is the trivial sensitivity vector.  The covariance matrix is the sum of an invertible diagonal matrix and one of rank one---see \eqref{cov}---, so that the inverse is obtained using the Woodbury matrix identity: 
\begin{align}
    \invcov{ij} &= \frac{\delta_{ij}}{\dev_i^2}
         - \frac{\sysdev^2\data_i\data_j}
                {\dev_i^2\dev_j^2\Big(1 + 
            \sysdev^2\sum\limits_{k} \frac{\data_k^2}{\dev_k^2}\Big)},
            \label{eq:invcov}\\
  \intertext{where we have introduced the statistical (uncorrelated) component of the uncertainty on the calibrated visibilities}
  \dev_i     &= \stadev\data_i \label{eq:devi}.
\end{align}

\correction{Appendix~\ref{ap:mu} shows the analytical derivation for the least squares estimate $\ppp{\mod}$ using the previous formulae. I write it in a way that highlights the generalisation of Eq.~(\ref{eq:mu2}) of the previous section:}
\begin{equation}
  \ppp{\mod} = 
    %\frac { 
    %  \sum\limits_i \frac{\data_i}{\dev_i^2}
    %}{ 
    %  \sum\limits_i 
    %      \Big( 1   
    %         + \corr\reldev^2\sum\limits_{j < i} \frac{(\data_i-\data_j)^2}{\dev_j^2}
    %       \Big)
    %      \frac{1}{\dev_i^2} 
    %}.
    \frac{\sum\limits_{i} \frac{\data_i}{\dev_i^2}}{\sum\limits_i \frac{1}{\dev_i^2}}
    \left( 1 + \sysdev^2\,\frac
            {\sum\limits_{i<j} \frac{(\data_i-\data_j)^2}{\dev_i^2\dev_j^2}} 
            {\sum\limits_i \frac{1}{\dev_i^2}}
        \right)^{-1}.
    \label{eq:mu}
\end{equation}

For small enough errors ($\error_i \ll \datamean$) the second-order Taylor development in $\error_i = \data_i - \datamean$ yields \correction{(see Appendix~\ref{ap:mutaylor})}: 
\begin{equation}
    %\ppp{\mod} \approx  \datamean 
    %              +\frac1n {\sum\limits_i \error_i}
    %                   - \frac {[2 + (n - 2)\corr]\reldev^2\datamean^2
    %                      }{
    %                       2n^2 \devmean^2             
    %                      }
    %                       \sum\limits_{i\ne j} (\error_i-\error_j)^2,
    \ppp{\mod} \approx 
        \datamean
      + \frac1n {\sum\limits_i \error_i}
      - \frac 1\datamean 
        \left(\frac1{2n}\frac{\sysdev^2}{\stadev^2}  + \frac 1{n^2}\right) 
      \sum\limits_{i\ne j} (\error_i-\error_j)^2.
    \label{eq:mutaylor}
\end{equation}

Since $\expect{\error_i} = 0$ and $\expect{(\error_i-\error_j)^2} = 2\stadev^2\datamean^2$, the \correction{expected value}
\begin{equation}
    %\expect{\ppp{\mod}} \approx \datamean 
    %        \left[ 1 -  
    %         \left(1-\frac1n\right)
    %         \left(2 + (n-2)\corr\right)
    %          \reldev^2 \right]
    \expect{\ppp{\mod}}  \approx \datamean 
            \left[ 1 -  
               \left(1-\frac1n\right)
               \left(2\stadev^2 + n\sysdev^2\right)
               \right]
    \label{eq:mutaylor2}
\end{equation}
is biased.  If the data are not correlated ($\sysdev = 0$), the bias is small
($\stadev^2$ to $2\stadev^2$) but it becomes larger for correlated data if the number of points is large ($n/2$ to $n \times \reldev^2$ for fully correlated data) as \citet{DAG94} already noted. This analytical derivation confirms the numerical simulation by \citet[][see their Fig.~4]{NEU12}. \correction{For visualisation purposes, Figure~\ref{fig:uncorr-peelle} shows a similar simulation of the bias as a function of the normalisation uncertainty $\reldev$ for various data sizes ($n = 2$ to 100). I have verified that it reproduces the quadratic behaviour of Eq.~(\ref{eq:mutaylor2}) for small values of $\stadev$ and $\sysdev$ (bias inferior to 10 to 20\% of \datamean).}

\correction{The bias from PPP arises, intuitively, because the modelled uncertainty is a non-constant function of the measured value. In the present case, data that fall below the average are given a lower uncertainty and, thus, a higher weight in the least squares fit. Conversely, data that fall above the average have a higher uncertainty and a lower weight. This fundamentally biases the estimate towards lower values.  The effect is much stronger with correlations because it impacts the $\sim n^2/2$ independent elements of the covariance matrix instead of being restricted to the $n$ diagonal ones.  In the literature, the puzzle is generally discussed as arising from a normalisation, as it it where it has been first identified. However, I show in appendix~\ref{ap:photon} that it is not necessary and determine the bias in the case of correlated photon noise.}

For spectro-interferometric observations with 4 telescopes, the number of correlated points can be over 1,000, so even with a low correlation coefficient, the \correction{bias can be significant}. For instance, a single GRAVITY observation in \correction{medium} spectral resolution yields $n = 6 \times 210$ visibility amplitudes. With an observed correlation of $\corr \approx 16$\% in the instrumental visibility amplitudes \citep{KAM20} and a typical $\reldev = 1$--2\% normalisation error, the bias on the calibrated visibilities could be 2--8\%.

@@

\begin{figure}
\centering
\subfigure[Under-resolved, linear least squares]{\includegraphics[width=\linewidth]{pdf/fit-example-quadratic.pdf}\label{fig:fitexample}}\\
\subfigure[Well resolved, non-linear least squares]{\includegraphics[width=\linewidth]{pdf/fit-example-gauss.pdf}\label{fig:fitexample2}}
\caption{\correction{Model fitting to simulated correlated data from a four-telescope interferometer (6 baselines) with medium spectral resolution (R = 100) with 2\% uncorrelated measurement error and 3\% correlated normalisation error} (light gray points with the measurement error bar).  \emph{Top:} Simulated under-resolved data $\data = 1-x^2$ (thick gray line) are fitted with linear least-squares model $\mod = a-bx^2$ using the four prescriptions for the covariance matrix. \emph{Bottom:} The same for well-resolved data $\data = \exp -3x^2$ and non-linear least-squares with model $\mod = a\exp -bx^2$.}
\end{figure}
